---
title: "cs455-TP analysis"
author: "Team 16: Zach Walsh, Erik West, Trevor Isaacson"
date: "4/24/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(455)
library(dplyr)
library(tidyverse)
library(Metrics)
library(GGally)
library(tree)
library(randomForest)
library(gbm)
library(ggplot2)
```

## Data Preview
```{r echo = FALSE}
final = read.csv("final.csv")
colnames(final) = c("State", "Week", "Pollution", "Deaths", "Powerplants", "TransmissionLines")
head(final)
```

Goal is to see if powerplants and transmission lines led to increased pollution in a state.  
If there is increased pollution, does it lead to more deaths in the state?

## Variable Analysis
```{r warning = FALSE, message=FALSE}
final %>% 
  group_by(State) %>%
  summarise(AvgPollution = sum(Pollution) / n()) %>%
  arrange(State) %>%
  ggplot() +
    geom_point(aes(x = State, y = AvgPollution)) +
    theme(text = element_text(size=10), axis.text.x = element_text(angle=90, hjust=1)) +
    ggtitle("Average Pollution Rating per State")

final %>%
  group_by(State) %>%
  summarise(Total = sum(Powerplants) / n()) %>%
  arrange(State) %>%
  ggplot() +
    geom_point(aes(x = State, y = Total)) +
    theme(text = element_text(size=10), axis.text.x = element_text(angle=90, hjust=1)) +
    ggtitle("Powerplants per State")

ggcorr(final)

summary(final)
```

## Split into Training / Testing:
```{r}
trn = sample(seq_len(nrow(final)), nrow(final)*0.75)
training = final[trn, ]
testing = final[-trn, ]
```

## Train and Test Various Regression and ML models to determine correlation of pollution with other variables
Multiple Linear Regression to determine if pollution is caused by powerplants and electrical lines
```{r}
linreg_PT = lm(Pollution ~ Powerplants + TransmissionLines, data = training)
summary(linreg_PT)
linreg_PT_predict = predict(linreg_PT, newdata = testing, type = "response")
RMSE_linreg_PT = sqrt(mean((testing$Pollution - linreg_PT_predict)^2))
print(paste("RMSE of linreg_PT on Testing Set: ", round(RMSE_linreg_PT, 4)))
```

Multiple Linear Regression to determine if pollution is caused by powerplants, electrical lines and state
```{r warning = FALSE}
linreg_PTS = lm(Pollution ~ Powerplants + TransmissionLines + State, data = training)
summary(linreg_PTS)
linreg_PTS_predict = predict(linreg_PTS, newdata = testing, type = "response")
RMSE_linreg_PTS = rmse(testing$Pollution, linreg_PTS_predict)
print(paste("RMSE of linreg_PTS on Testing Set: ", round(RMSE_linreg_PTS, 4)))
```

Multiple Linear Regression to determine if pollution is caused by powerplants, electrical lines, state, week
```{r warning = FALSE}
linreg_PTSW = lm(Pollution ~ Powerplants + TransmissionLines + State + Week, data = training)
summary(linreg_PTSW)
linreg_PTSW_predict = predict(linreg_PTSW, newdata = testing, type = "response")
RMSE_linreg_PTSW = rmse(testing$Pollution, linreg_PTSW_predict)
print(paste("RMSE of linreg_PTSW on Testing Set: ", round(RMSE_linreg_PTSW, 4)))
```

Fit Pollution Tree with Powerplants, Electrical Lines and Week
```{r} 
# fit regression tree
tree_PTW = tree(Pollution ~ Powerplants + TransmissionLines + Week, data = training)

tree_PTW_predict = predict(tree_PTW, newdata = testing)
RMSE_tree_PTW = rmse(testing$Pollution, tree_PTW_predict)
print(paste("Test RMSE of Tree_PTS: ", round(RMSE_tree_PTW, 4)))

# cross validation
tree_PTW_cv = cv.tree(tree_PTW)
ggplot() +
  geom_point(aes(x = tree_PTW_cv$size, y = tree_PTW_cv$dev)) + 
  geom_line(aes(x = tree_PTW_cv$size, y = tree_PTW_cv$dev)) + 
  ylab("CV Error Rate") + xlab("Size") + ggtitle("CV Classification Error Rate")

# pruned by CV
prune_tree_PTW <- prune.tree(tree_PTW, best = 7)
plot(tree_PTW)
text(tree_PTW, cex = 0.65, digits = 4, pretty = 0)
title("Powerplants, Lines and Week")
```

Random Forest for Pollution on All Variables
```{r}
bag_pollution <- randomForest(Pollution ~ State + Week + TransmissionLines + Powerplants, data = training, importance = TRUE)

bag_pollution_predict = predict(bag_pollution, testing, type = "response")
bag_pollution_RMSE = rmse(testing$Pollution, bag_pollution_predict)
print(paste("Test RMSE of bag_pollution: ", round(bag_pollution_RMSE, 4)))

ggplot() + 
  geom_point(aes(x = bag_pollution_predict, y = testing$Pollution), alpha = 0.25) + 
  geom_abline(slope = 1, intercept = 0) + 
  ggtitle("Performance of bag_pollution on Test Set")

data.frame(bag_pollution$importance) %>%
  mutate(variable = rownames(bag_pollution$importance)) %>%
  mutate(variable = factor(variable, levels = variable[order(X.IncMSE)])) %>%
  ggplot() + ggtitle("Predictive Power") + xlab("% Decreasing MSE") + 
  geom_point(aes(X.IncMSE, variable))  

data.frame(bag_pollution$importance) %>%
  mutate(variable = rownames(bag_pollution$importance)) %>%
  mutate(variable = factor(variable, levels = variable[order(IncNodePurity)])) %>%
  ggplot() + ggtitle("Predictive Power") + xlab("Node Purity") + 
  geom_point(aes(IncNodePurity, variable))  
```
Boosting for Pollution with All Variables
```{r}
lambdas = seq(0, 5, length.out = 100)
trainErrors = rep(NA, length(lambdas))
testErrors = rep(NA, length(lambdas))
for(i in 1:length(lambdas)){
  boost = gbm(Pollution ~ State + Week + Powerplants + TransmissionLines, data = training, n.trees = 100, shrinkage = lambdas[i], distribution = "gaussian")
  trainPred = predict(boost, training, n.trees = 100)
  testPred = predict(boost, testing, n.trees = 100)
  trainErrors[i] = mean((training$Pollution - trainPred)^2)
  testErrors[i] = mean((testing$Pollution - testPred)^2)
}

boostedPollution = gbm(Pollution ~ State + Week + Powerplants + TransmissionLines, data = training, n.trees = 500, shrinkage = lambdas[which.min(testErrors)], distribution = "gaussian")
summary(boostedPollution)

boostedPollution_pred = predict(boostedPollution, testing)
boostedPollution_RMSE = round(sqrt(mean((testing$Pollution - boostedPollution_pred)^2)), 4)
print(paste("Testing MSE for Boosted Pollution Model:", boostedPollution_RMSE))
```


## Train and Test Various Regression and ML models to determine correlation of state deaths with other variables
Multiple Linear Regression for Death with All Variables
```{r warning = FALSE}
linreg_DA = lm(Deaths ~ ., data = training)
summary(linreg_DA)
linreg_DA_predict = predict(linreg_DA, newdata = testing, type = "response")
RMSE_linreg_DA = sqrt(mean((testing$Deaths - linreg_DA_predict)^2))
print(paste("RMSE of linreg_PT Testing Set: ", round(RMSE_linreg_DA, 4)))
```


Random Forest for Deaths on All Variables
```{r}
bag_deaths <- randomForest(Deaths ~ ., data = training, importance = TRUE)

bag_deaths_predict = predict(bag_deaths, testing, type = "response")
bag_deaths_RMSE = rmse(testing$Deaths, bag_deaths_predict)
print(paste("Test RMSE of bag_pollution: ", round(bag_deaths_RMSE, 4)))

ggplot() + 
  geom_point(aes(x = bag_deaths_predict, y = testing$Deaths), alpha = 0.25) + 
  geom_abline(slope = 1, intercept = 0) + 
  ggtitle("Performance of Bagging on Testing Set")

data.frame(bag_deaths$importance) %>%
  mutate(variable = rownames(bag_deaths$importance)) %>%
  mutate(variable = factor(variable, levels = variable[order(X.IncMSE)])) %>%
  ggplot() + ggtitle("Predictive Power") + xlab("% Decreasing MSE") + 
  geom_point(aes(X.IncMSE, variable))  

data.frame(bag_deaths$importance) %>%
  mutate(variable = rownames(bag_deaths$importance)) %>%
  mutate(variable = factor(variable, levels = variable[order(IncNodePurity)])) %>%
  ggplot() + ggtitle("Predictive Power") + xlab("Node Purity") + 
  geom_point(aes(IncNodePurity, variable))  
```

Random Forest for Deaths on All Variables Except State
```{r}
bag_DES <- randomForest(Deaths ~ Week + TransmissionLines + Powerplants + Pollution, data = training, importance = TRUE)

bag_DES_predict = predict(bag_DES, testing, type = "response")
bag_DES_RMSE = rmse(testing$Deaths, bag_DES_predict)
print(paste("Test RMSE of bag_deaths without states: ", round(bag_DES_RMSE, 4)))

ggplot() + 
  geom_point(aes(x = bag_DES_predict, y = testing$Deaths), alpha = 0.25) + 
  geom_abline(slope = 1, intercept = 0) + 
  ggtitle("Performance of Bagging on Test Set")

data.frame(bag_DES$importance) %>%
  mutate(variable = rownames(bag_DES$importance)) %>%
  mutate(variable = factor(variable, levels = variable[order(X.IncMSE)])) %>%
  ggplot() + ggtitle("Predictive Power") + xlab("% Decreasing MSE") + 
  geom_point(aes(X.IncMSE, variable))  

data.frame(bag_DES$importance) %>%
  mutate(variable = rownames(bag_DES$importance)) %>%
  mutate(variable = factor(variable, levels = variable[order(IncNodePurity)])) %>%
  ggplot() + ggtitle("Predictive Power") + xlab("Node Purity") + 
  geom_point(aes(IncNodePurity, variable))  
```

Boosting for Death with All Variables
```{r}
lambdas = seq(0, 5, length.out = 100)
trainErrors = rep(NA, length(lambdas))
testErrors = rep(NA, length(lambdas))
for(i in 1:length(lambdas)){
  boost = gbm(Deaths ~ ., data = training, n.trees = 100, shrinkage = lambdas[i], distribution = "gaussian")
  trainPred = predict(boost, training, n.trees = 100)
  testPred = predict(boost, testing, n.trees = 100)
  trainErrors[i] = mean((training$Deaths - trainPred)^2)
  testErrors[i] = mean((testing$Deaths - testPred)^2)
}

boosted = gbm(Deaths ~ ., data = training, n.trees = 500, shrinkage = lambdas[which.min(testErrors)], distribution = "gaussian")
summary(boosted)

boosted_pred = predict(boosted, testing)
boosted_RMSE = round(sqrt(mean((testing$Deaths - boosted_pred)^2)), 4)
print(paste("Testing MSE for Boosted Model:", boosted_RMSE))
```


```{r}
ggplot() + 
  geom_point(aes(x = boosted_pred, y = testing$Deaths), alpha = 0.25) + 
  geom_abline(slope = 1, intercept = 0) + 
  ggtitle("Performance of Boosting Deaths on Test Set")

```


## RMSE Pollution Table
```{r}
method = c("Linear Regression Pollution (PP + TL)", "Linear Regression Pollution (PP + TL + St)", "Linear Regression Pollution (All)", "Trees (No State)", 
           "Bagging Pollution RMSE (All)", "Boosting Pollution RMSE (All)")
RMSEpollution = c(RMSE_linreg_PT, RMSE_linreg_PTS, RMSE_linreg_PTSW, RMSE_tree_PTW, bag_pollution_RMSE, boostedPollution_RMSE)
maxPollution = max(final$Pollution)
minPollution = min(final$Pollution)
normRMSEPollution = c(RMSE_linreg_PT/(maxPollution-minPollution), RMSE_linreg_PTS/(maxPollution-minPollution), RMSE_linreg_PTSW/(maxPollution-minPollution), RMSE_tree_PTW/(maxPollution-minPollution), bag_pollution_RMSE/(maxPollution-minPollution), boostedPollution_RMSE/(maxPollution-minPollution))

finalRMSEpollution = data.frame(Methods = method, RMSE = round(RMSEpollution,4), NormRMSE = round(normRMSEPollution, 4))
finalRMSEpollution
```

## RMSE Death Table
Normalizing RMSE scores gives us a better understanding of how "good" our model does at predicting
Values are 0 to 1
Closer to 0 represents a better number
```{r}
method = c("Linear Regression Deaths (All)", "Bagging Deaths (All)", "Bagging Deaths (No States)", "Boosting Death (All)")
RMSEdeaths = c(RMSE_linreg_DA, bag_deaths_RMSE, bag_DES_RMSE, boosted_RMSE)
maxDeaths = max(final$Deaths)
minDeaths = min(final$Deaths)
normRMSEDeaths = c(RMSE_linreg_DA/(maxDeaths-minDeaths), bag_deaths_RMSE/(maxDeaths-minDeaths), bag_DES_RMSE/(maxDeaths-minDeaths), boosted_RMSE/(maxDeaths-minDeaths))

finalRMSEdeaths = data.frame(Methods = method, RMSE = round(RMSEdeaths, 4), NormRMSE = round(normRMSEDeaths,4))
finalRMSEdeaths
```


## Final Model: Does higher pollution correlate to more deaths?
```{r message = FALSE}
lambdas = seq(0, 5, length.out = 100)
trainErrors = rep(NA, length(lambdas))
testErrors = rep(NA, length(lambdas))
for(i in 1:length(lambdas)){
  boost = gbm(Deaths ~ Pollution, data = training, n.trees = 100, shrinkage = lambdas[i], distribution = "gaussian")
  trainPred = predict(boost, training, n.trees = 100)
  testPred = predict(boost, testing, n.trees = 100)
  trainErrors[i] = mean((training$Deaths - trainPred)^2)
  testErrors[i] = mean((testing$Deaths - testPred)^2)
}

finalBoosted = gbm(Deaths ~ Pollution, data = training, n.trees = 500, shrinkage = lambdas[which.min(testErrors)], distribution = "gaussian")

finalBoosted_pred = predict(finalBoosted, testing)
finalBoosted_RMSE = round(sqrt(mean((testing$Deaths - finalBoosted_pred)^2)), 4)
print(paste("Testing MSE for Boosted Model on Test Set:", finalBoosted_RMSE))
```



